Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package \vskip15pt Supplementary Materials

### 2018-06-18 docs/MEE_submission/manuscript_MEE.tex
MS revisions

\item Here we present the R package \pkg{metaDigitise} which extracts descriptive statistics such as means, standard deviations and correlations from the four plot types: 1) mean/error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. \pkg{metaDigitise} is user-friendly and easy to learn as it interactively guides the user through the data extraction process. Notably, it enables large-scale extraction by automatically loading image files, letting the user stop processing, edit and add to the resulting data fame at any point.

@@

In many different contexts, researchers make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Glass1976,Koricheva2013,Naka2017}. Meta-analysis relies on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature that have been reported in the text or tables of research papers. Descriptive statistics are also, however, frequently presented in figures and so need to be manually extracted using digitising programs. While inferential statistics (e.g., \textit{t}- and \textit{F}-statistics) are often presented along side descriptive statistics, and can be used to derive effect sizes, descriptive statistics are much more appropriate to use because sources of non-independence in experimental designs can be dealt with more easily \citep{Noble2017}.

Although there are several tools that data extraction from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}), these tools do not cater to needs of meta-analysis for three main reasons. First, they typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently a large amount of downstream data manipulation is required, that is different across plots types. For example, data are frequently presented in mean/error plots (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. Error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Finally, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software.

Data extraction from figures is therefore an incredibly time-consuming process as existing software does not provide an optimized research pipeline to facilitate data extraction and editing. Furthermore, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}.

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or use in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations on figures, make corrections and access calibration and proceeded data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{Directory Structure and Reproducibility}
The \pkg{metaDigitise} package was created with the idea that users would have multiple images to extract from and therefore operates in the same way whether the user has one or multiple images. There is one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures. \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. The user should think carefully about their directory structure early on in their project. Although different directory structures may be used, we would recommend having all files for one project in a single directory with an informative and unambiguous naming scheme for images to help identify the paper and figure that data come from (e.g. paper\_figure\_trait.png).

\fct{metaDigitise} recognizes all the images in a directory and automatically imports them one by one, allowing the user to extract the relevant information about a figure as they go. Having all figures in one directory therefore expedites digitisation by preventing users from having to constantly change directories and / or open new images. The data from each completed image is automatically saved as a \code{metaDigitise} object in a separate .RDS file to a \code{caldat} directory that is created within the parent directory when first executing \fct{metaDigitise}. These files enable re-plotting and editing of images at a later point (see below). When run, \fct{metaDigitise} also identifies the images within a directory that have been previously digitised and only imports new images to process. The data of all images is then automatically integrated into the final output. This means that all figures do not need to be extracted at one time and new figures can be added to the directory as the project develops.

@@

\section{Image Processing}
Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data'. Selecting `Process New Images' starts the digitisation process on images within the directory that have not preciously been digitised; the other functions are discussed below.

@@

\subsection{Adding Sample Sizes to Previous Digitisations}
In many cases sample sizes may not be readily available when digitising figures. This information does not need to be added a the time of digitisation. To expedite finding and adding these sample sizes at a later point, \fct{metaDigitise} has a specific edit option that allows users to enter previously omitted sample sizes. This first identifies missing sample sizes in the digitised output, re-plots the relevant figures and prompts the user to enter the sample sizes for the relevant groups in the figure, one by one.

@@

\section{Software Validation}
In order to evaluate the consistency of digitisation with \pkg{metaDigitise} between users, we got 14 people to digitise the same set of 14 figures created from a simulated dataset (see supplements). We found no evidence for any inter-observer variability in digitisations for the mean (ICC = 0, 95\% CI = 0 to 0.029, \textit{p} = 1), SD (ICC = 0, 95\% CI = 0 to 0.033, \textit{p} = 0.5) or correlation coefficient (ICC = 0.053, 95\% CI = 0 to 0.296, \textit{p} = 0.377). There was little bias between digitised and true values, on average 1.63\% (mean = 0.02\%, SD = 4.9\%, \textit{r} = -0.03\%) and there were small absolute differences between digitised and true values, on average 2.18\% (mean = 0.40\%, SD = 5.81\%, \textit{r} = 0.33\%) across all three summary statistics. SD estimates from digitisations are clearly most error prone. The mean absolute differences for each plot type clearly show that this effect is driven by extraction from box plots and histograms (\% difference; box plot: 15.805, histogram: 5.210, mean/error: 1.500, scatter plot: 0.433). SD estimation from box plot summary statistics is known to be more error prone, especially at small sample sizes \citep{Wan2014}.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Line extraction is not particularly useful for most meta-analyses, although we recognise that it may be useful in other fields. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

>>>>

\item Here we present the R package \pkg{metaDigitise} which extracts descriptive statistics such as means, standard deviations and correlations from the four plot types: 1) mean/error plots (e.g. bar graphs with standard errors), 2) box plots, 3) scatter plots and 4) histograms. \pkg{metaDigitise} is user-friendly and easy to learn as it interactively guides the user through the data extraction process. Notably, it enables large-scale extraction by automatically loading image files, letting the user stop processing, edit and add to the resulting data-frame at any point.

@@

%%Most notably, this includes meta-analysis, which is increasingly common in ecology and evolution. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects \citep{Glass1976,Koricheva2013,Naka2017}.

%% While inferential statistics (e.g., \textit{t}- and \textit{F}-statistics) are often presented along side descriptive statistics, and can be used to derive effect sizes, descriptive statistics are much more appropriate to use because sources of non-independence in experimental designs can be dealt with more easily \citep{Noble2017}.

In many different contexts, researchers make use of data presented in primary literature. In the ecology and evolution, this most notably includes comparative and meta-analyses. These techniques rely on descriptive statistics (e.g. means, standard deviations (SD), sample sizes, correlation coefficients) extracted from primary literature. As well as being presented in the text or tables of research papers, descriptive statistics are frequently presented in figures and so need to be manually extracted using digitising programs.

Although there are several tools that extract data from figures (e.g. \proglang{DataThief} \citep{DataThief}, \proglang{GraphClick} \citep{GraphClick}, \proglang{WebPlotDigitizer} \citep{WebPlotDigitizer}), these tools do not cater to needs of meta-analysis for four main reasons (here we focus on meta-analysis, although many  points apply to extraction for comparative analysis). First, although meta-analysis is an important tool in consolidating the data from multiple studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending or replicating studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and aid in resolving the reproducibility crises seen in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, Parker2016}. Second, digitising programs do not allow the integration of metadata at the time of data extraction, such as experimental group or variable names, and sample sizes. This makes the downstream calculations laborious, as information has to be added later using different software. Third, existing programs do not import sets of images for the user to systematically work through. Instead they require the user to manually import images one by one, and export data into individual files, that need to be imported and edited using different software. Finally, digitising programs typically only provide the user with calibrated \textit{x,y} coordinates from imported figures, and do not differentiate between common plot types that are used to present data. Consequently a large amount of additional data manipulation is required, that is different across plots types. For example, data are frequently presented in plots with means and standard errors or confidence intervals (Figure \ref{fig:all_extract}A), from which the user wants a mean and SD for each group presented. From \textit{x,y} coordinates, users must manually discern between mean and error coordinates and assign points to groups. The error then needs to be calculated as the deviation from the mean, and then transformed to SD, according to the type of error presented.

Data extraction from figures is therefore an incredibly time-consuming process as existing software does not provide an optimized, reproducible research pipeline to facilitate data extraction and editing. Here, we present an interactive \proglang{R} package, \pkg{metaDigitise} (available at https://github.com/daniel1noble/metaDigitise), which is designed for large scale, reproducible data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools to extract data from common plot types (mean/error plots, box plots, scatter plots and histograms, see Figure \ref{fig:all_extract}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. The necessary calculations are carried out on calibrated data immediately after extraction so that comparable summary statistics can be obtained quickly. Summary data from multiple figures is returned into a single data frame which can be can easily exported or use in downstream analysis within \proglang{R}. Completed digitisations are automatically saved for each figure, meaning users can redraw their digitisations on figures, make corrections and access calibration and proceeded data. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analyses to be updated more efficiently.

@@

\section{\pkg{metaDigitise} and Reproducibility}
The \pkg{metaDigitise} package has one main function, \fct{metaDigitise}, which interactively takes the user through the process of extracting data from figures. \fct{metaDigitise} works on a directory containing images of figures copied from primary literature, in .png, .jpg, .tiff, .pdf format, specified to \fct{metaDigitise} through the \code{dir} argument. \fct{metaDigitise} recognizes all the images in the given directory and automatically imports them one by one, allowing the user to extract the relevant information about a figure as they go. It is worth the user thinking carefully about their directory structure early on in their project. Although different directory structures may be used, we would recommend having all figures for one project in a single directory with an informative and unambiguous naming scheme (e.g. paper\_figure\_trait.png). This expedites digitisation by preventing users from having to constantly change directories and / or open new images.

The data from each completed image is automatically saved as a \code{metaDigitise} object in a separate .RDS file to a \code{caldat} directory that is created within the parent directory when first executing \fct{metaDigitise}. These files enable re-plotting and editing of images at a later point (see below). When run, \fct{metaDigitise} also identifies the images within a directory that have been previously digitised and only imports new images to process. The data of all images is then automatically integrated into the final output. This means that all figures do not need to be extracted at one time and new figures can be added to the directory as the project develops.

@@

\section{Image Processing}
Running \fct{metaDigitise} presents the user with three options; `Process new images', `Import existing data' or `Edit existing data'. Selecting `Process New Images' starts the digitisation process on images within the directory that have not previously been digitised; the other functions are discussed below.

@@

\subsection{Adding Sample Sizes to Previous Digitisations}
In many cases sample sizes may not be readily available when digitising figures. This information does not need to be added at the time of digitisation. To expedite finding and adding these sample sizes at a later point, \fct{metaDigitise} has a specific edit option that allows users to enter previously omitted sample sizes. This first identifies missing sample sizes in the digitised output, re-plots the relevant figures and prompts the user to enter the sample sizes for the relevant groups in the figure, one by one.

@@

\section{Software Validation}
In order to evaluate the consistency of digitisation with \pkg{metaDigitise} between users, fourteen people digitized sets of 14 identical images created from a simulated dataset (see supplements). We found no evidence for any inter-observer variability in digitisations for the mean (ICC = 0, 95\% CI = 0 to 0.029, \textit{p} = 1), SD (ICC = 0, 95\% CI = 0 to 0.033, \textit{p} = 0.5) or correlation coefficient (ICC = 0.053, 95\% CI = 0 to 0.296, \textit{p} = 0.377). There was little bias between digitised and true values, on average 1.63\% (mean = 0.02\%, SD = 4.9\%, \textit{r} = -0.03\%) and there were small absolute differences between digitised and true values, on average 2.18\% (mean = 0.40\%, SD = 5.81\%, \textit{r} = 0.33\%) across all three summary statistics. SD estimates from digitisations are clearly most error prone. The mean absolute differences for each plot type clearly show that this effect is driven by extraction from box plots and histograms (\% difference; box plot: 15.805, histogram: 5.210, mean/error: 1.500, scatter plot: 0.433). SD estimation from box plot summary statistics is known to be more error prone, especially at small sample sizes \citep{Wan2014}.

@@

In contrast to some other packages \pkg{metaDigitise} does not extract lines from figures. Line extraction is not particularly useful for most comparative or meta-analytic work, although we recognise that it may be useful in fields other than these. Should a user like to extract lines with \pkg{metaDigitise}, we would recommend extracting data as a scatter plot, and clicking along the line in question. A model can then be fitted to these points (accessed by choosing to return calibrated rather than summary data) to estimate the parameters needed.

### 2018-06-18 docs/MEE_submission/supplements.tex
MS revisions

\section{Comparison with existing software}
\begin{table}[h!]
\centering
\rotatebox{90}{%
\begin{minipage}{1.35\textwidth}
\small
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean/error plots      & \checkmark   & \checkmark & \checkmark & $\times$   & $\times$       & \checkmark$^7$ & $\times$    \\
Boxplots                  & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Histograms                & \checkmark   & $\times$   & $\times$   & $\times$   & \checkmark$^7$ & $\times$       & $\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark & \checkmark     & $\times$       & $\times$    \\
Entry of metadata         & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Summarising data          & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Multiple image processing & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Reproducable$^9$  & \checkmark   & \checkmark & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Automated point detection & $\times$     & \checkmark &     $\times$      & \checkmark & \checkmark     & \checkmark     & $\times$    \\
Line extraction           & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Zoom                      & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Log axis                  & $\times$   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Dates                     & $\times$   & $\times$   & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Asymmetric error bars     & $\times$     & $\times$   & \checkmark & $\times$   & $\times$       & $\times$       & $\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline

>>>>

\section{Comparison with existing software}
\begin{table}[h!]
\centering
\rotatebox{90}{%
\begin{minipage}{1.35\textwidth}
\small
\begin{tabular}{lccccccc}
\hline
Function                  & metaDigitise & GraphClick$^1$ & DataThief$^2$  & DigitizeIt$^3$ & WebPlotDigitizer$^4$ & metagear$^5$ & digitize$^6$\\
\hline
Scatterplots              & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & \checkmark$^7$ & \checkmark \\
Mean/error plots      & \checkmark   & \checkmark & \checkmark & $\times$   & $\times$       & \checkmark$^7$ & $\times$    \\
Boxplots                  & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Histograms                & \checkmark   & $\times$   & $\times$   & $\times$   & \checkmark$^7$ & $\times$       & $\times$    \\
Graph rotation$^8$        & \checkmark   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Groups                    & \checkmark   & \checkmark & $\times$   & \checkmark & \checkmark     & $\times$       & $\times$    \\
Entry of metadata         & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Summarising data          & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Multiple image processing & \checkmark   & $\times$   & $\times$   & $\times$   & $\times$       & $\times$       & $\times$    \\
Reproducable$^9$  & \checkmark   & \checkmark & \checkmark & $\times$   & \checkmark     & \checkmark       & $\times$    \\
Automated point detection & $\times$     & \checkmark &     $\times$      & \checkmark & \checkmark     & \checkmark     & $\times$    \\
Line extraction           & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Zoom                      & $\times$     & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Log axis                  & $\times$   & \checkmark & \checkmark & \checkmark & \checkmark     & $\times$       & $\times$    \\
Dates                     & $\times$   & $\times$   & \checkmark & $\times$   & \checkmark     & $\times$       & $\times$    \\
Asymmetric error bars     & $\times$     & $\times$   & \checkmark & $\times$   & $\times$       & $\times$       & $\times$    \\
Freeware                  & \checkmark$^{10}$ & \checkmark$^{11}$ & \checkmark$^{11}$ & $\times$$^{11}$ & \checkmark$^{11}$ & \checkmark$^{10}$ & \checkmark$^{10}$\\
\hline
